---
title: "Geography 176A"
author: "[Wesley Noble](https://wes-noble.github.io/)"
date: "08-14-2020"
subtitle: 'Lab 02: COVID-19 Pandemic'
output:
  html_document: 
    theme: journal
---

#My Project

##Question 1

```{r, messages = FALSE, echo = FALSE}

library(tidyverse)
library(knitr)
library(zoo)
library(readxl)

```

```{r, messages = FALSE}

covid = read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv")

PopulationEstimates = read_excel("/users/noblex/github/geog-176A-labs/data/PopulationEstimates.xls", skip = 2)

#Question 1

state.of.interest = "California"

(state_covid = covid %>%
  filter(state == state.of.interest) %>%
  group_by(county) %>%
  mutate(newCases = cases - lag(cases)) %>%
  ungroup())

(county_most_new_cases = state_covid %>%
    filter(date == max(date)) %>%
    slice_max(newCases, n = 5) %>%
    select(county, newCases))

(county_most_cumulative_cases = state_covid %>%
    filter(date == max(date)) %>%
    slice_max(cases, n = 5) %>%
    select(county, cases))

kable(county_most_new_cases, caption = "Most New Cases California Counties",
      col.names = c("County", "New Cases"),
      format.args = list(big.mark = ","))

kable(county_most_cumulative_cases, caption = "Most Cumulative Cases California Counties",
      col.names = c("County", "Total Cases"),
      format.args = list(big.mark = ","))

(pop_by_county = PopulationEstimates %>%
    filter(State == "CA") %>%
    select(pop19 = POP_ESTIMATE_2019, state = State, county = Area_Name, fips = FIPStxt) %>%
    group_by(county) %>%
    slice_max(pop19, n=1))

(covid_cum_pop = right_join(pop_by_county, state_covid, by = "fips") %>%
    filter(date == max(date)) %>%
    mutate(cases_p_cap = (cases/pop19)) %>%
    slice_max(cases_p_cap, n = 5) %>%
    select(county.y, cases_p_cap))

(covid_newcases_pop = right_join(pop_by_county, state_covid, by = "fips") %>%
    filter(date == max(date)) %>%
    mutate(newcases_p_cap = (newCases/pop19)) %>%
    slice_max(newcases_p_cap, n = 5) %>%
    select(county.y, newcases_p_cap))

kable(covid_cum_pop, caption = "Most Cumulative Cases Per Capita California Counties",
      col.names = c("County", "Cases Per Capita"),
      format.args = list(big.mark = ","))

kable(covid_newcases_pop, caption = "Most Daily New Cases Per Capita California Counties",
      col.names = c("County", "Daily New Cases Per Capita"),
      format.args = list(big.mark = ","))

covid_pop = right_join(pop_by_county, state_covid, by = "fips")

(newCases_last14days_per100000 = covid_pop %>%
  filter(date > max(date) - 14, date < max(date)) %>%
  select(county = county.y, newCases, pop19, date) %>%
  group_by(county, pop19) %>%
  summarise(newCases = sum(newCases, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(pop_per_100000 = (pop19/100000), newCases_per100000 = (newCases/pop_per_100000)))


(total_state_cases = state_covid %>%
  filter(date == max(date)) %>%
  group_by(county) %>%
  summarise(cases = sum(cases, na.rm = TRUE)) %>%
  ungroup() %>%
  summarise(cases = sum(cases, na.rm = TRUE)) %>%
  pull(cases))


(total_state_newCases = state_covid %>%
  filter(date == max(date)) %>%
  group_by(county) %>%
  summarise(newCases = sum(newCases, na.rm = TRUE)) %>%
  ungroup() %>%
  summarise(newCases = sum(newCases, na.rm = TRUE)) %>%
  pull(newCases))


(healthy_counties = newCases_last14days_per100000 %>%
  filter(newCases_per100000 < 100) %>%
  pull(county))

```

---

There are `r total_state_cases` 
total cases in California, 
`r total_state_newCases ` 
new cases in California, and these are the counties that are deemed safe by the California Department of Public Health based on the number of new cases per 100,000 residents over the past 14 days: 
`r healthy_counties`. 


---

```{r}
#Question 2

data2 = covid %>%
  filter(state %in% c("California", "New York", "Louisiana", "Florida")) %>%
  group_by(state, date) %>%
  summarise(cases = sum(cases, na.rn = TRUE)) %>%
  ungroup() %>%
  group_by(state) %>%
  mutate(newCases = cases - lag(cases), roll7 = zoo::rollmean(newCases, 7, fill = NA, allign = "right")) %>%
  ungroup() %>% 
  filter(newCases > 0)

  ggplot(data = data2, aes(x = date, y = newCases)) +
  geom_col(aes(y = newCases), col = NA, fill = "tomato1") +
  geom_line(aes(y = roll7), col = "violetred4", size = 1) +
  facet_grid(~state, scales = "free_y")+
  theme_update() +
  labs(title = paste("New Reported Cases by Day in Select States"), caption = "Lab 2 - Question 2", x = "Date",
       y = "New Cases")

newCases_pop = PopulationEstimates %>%
  select(pop19 = "POP_ESTIMATE_2019", areaname = "Area_Name") %>%
  right_join(data2, by = c("areaname" = "state")) %>%
  mutate(cum_cases_percapita = (cases/pop19), newCases_percapita = (newCases/pop19), roll7 = zoo::rollmean(newCases_percapita, 7, fill = NA, allign = "right")) 
  

ggplot(data = newCases_pop, aes(x = date, y = newCases_percapita)) +
  geom_col(aes(y = newCases_percapita), col = NA, fill = "turquoise2") +
  geom_line(aes(y = roll7), col = "violetred4", size = 1) +
  facet_grid(~areaname, scales = "free_y")+
  theme_update() +
  labs(title = paste("New Reported Cases by Day in Select States Per Capita"), caption = "Lab 2 - Question 2", x = "Date",
       y = "New Cases Per Capita")

```

---

When scaling the daily new cases to the population of the respected states, we see that although California shows high daily new cases, it has a much smaller daily new cases per capita distribution. On the other hand, we observe Louisiana to have the lowest daily new cases in our comparison, but when estimated based on state population we find it to have a much greater and stark distribution of new cases per capita. New York and Florida roughly follow the same distribution between their daily new cases and daily new cases per capita. 

---



